\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{Assessing income inequality with tax data - Switzerland from 1917 to 2010}

### 2014-11-25 paper/paper.tex
major sync,word files, figures, bern tax data analyses

\usepackage[hmarginratio=1:1,top=32mm,columnsep=20pt]{geometry} % Document margins
\usepackage{multicol} % Used for the two-column layout of the document
\usepackage[hang, small,labelfont=bf,up,textfont=it,up]{caption} % Custom captions under/above floats in tables or figures
\usepackage{booktabs} % Horizontal rules in tables
\usepackage{float} % Required for tables and figures in the multi-column environment - they need to be placed in specific locations with the [H] (e.g. \begin{table}[H])
\usepackage{hyperref} % For hyperlinks in the PDF
\usepackage{graphicx}

@@

\usepackage{fancyhdr} % Headers and footers
\pagestyle{fancy} % All pages have headers and footers
\fancyhead{} % Blank out the default header
\fancyfoot{} % Blank out the default footer
\fancyhead[C]{Running title $\bullet$ November 2012 $\bullet$ Vol. XXI, No. 1} % Custom header text
\fancyfoot[RO,LE]{\thepage} % Custom footer text
%\bibliographystyle{plain}
\bibliographystyle{unsrtnat}

@@

% 3.Abschnitt: What do we know about Switzerland
% Offical Data Collections on development of inequality
As we focus our paper on the case of Switzerland, it is important to embed our work in the context of given publications concerning inequality in income. What is known about Switzerland so far? Looking for official data, three main sources has to be mentioned, which can be considered as de facto official data sources: EU-SILC, HBS and LIS-data. Figure XY shows the results stemming from this three sources while looking at Gini of equivalised disposable income. Up to the day, EU-SILC or Statistics on Income and Living Conditions is the main source used for policy monitoring at EU-level. The main focus of EU-SILC is to collect data on a common ``framework'' to ensure comparability among EU-countries and countries living around or within the EU. As a Non-EU member Switzerland implemented the instrument not from the beginning (2004) but as from 2007. Therefore this times-series doesn't cover time points before 2007. As graph XY shows, following the results from SILC income inequality decreased from 2007 to 2013.\footnote{Data shown in the graph was downloaded from the Eurostat Metadata-portal \url{http://epp.eurostat.ec.europa.eu/cache/ITY_SDDS/EN/ilc_esms.htm} last accessed 21.Mai 2014.} The second important source concerning the distribution of income is the Household Budget Survey (HBS). The main focus of this survey lays in providing detailed data on household budgets. This allows researcher to look at different income concepts like income before and after public transfers. Since 2000 the survey has been conducted on a continuous basis, which allows to look at a consistent time series from 2000 to 2011. As it can be seen from graph XY the trend is rather stable.\footnote{figures shown in the graph were calculated out of the original datasets, which were kindly provided by Swiss Federal Statistical Office.} Both time-series (SILC and HBS) cover a relatively short time period. A longer period is covered in the LIS-Data-set (1982-2004). Data-provider for the LIS Data is the Swiss Federal Statistical Office too. In contrast the the aforementioned surveys the LIS-data is harmonized out of three surveys: Swiss Income and Wealth Survey (1982), Swiss Poverty Survey (1992) and the Income and Consumption survey (2000, 2002, 2004).  All in all the LIS dataset contains the longest time series on inequality for Switzerland. Analyzing this data \citet{gornick_income_2013} found for Switzerland a quite substantially decreases in income inequality, contradictory to the development in most other western countries. This result is supported by \citet{grabka_evolution_2012} analyzing the Swiss Household Panel (2000-2009).\footnote{A further official database for income distribution is the OECD-Database. It includes measures from Income and Consumption survey as well. Additional data for 2008 is available from EU-Survey of Income and Living Conditions (EU-SILC). But this change in survey is considered as a strict break. Comparison before and since 2008 is not recommended (OECD 2012:315). For the sake of completeness the database construtcted by \citet{deininger_new_1996} and the World Income Inequality Database (WIID) have to be mentioned. Both datasets do not contain figures  for recent years for Switzerland. A further important database on inequality is the GINI database which has been derived from the GINI Country Reports \citep{nolan_changing_2014}. But this dataset doesn't cover Switzerland.}  \\

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/Swiss_inequality_Plot}

@@

\emph{Measuring inequality or concentration} \\
To be able to make qualifying statements about a distribution or to compare different distributions, the concept of inequality turned out to be the most appropriate and thus the most commonly used dimension. The Gini coefficient is the most known measure and mainly used for international comparison. As it is derived from the Lorenz-curve, the quantified amount of inequality can unpretentiously be described in a formal and visual way. Therefore the Gini coefficient is easily interpretable. Furthermore it has several desired statistical properties \citet{engelhardt_modelle_2000}. (1) ``\textbf{principle of population}'': the assessment of inequality is independent of the population size (2) ``\textbf{Requirement of Bresciani-Turroni}'': the measure is sensitive for changes of income shares, but not for absolute changes (e.g. doubling of all income) (3) ``\emph{weak principle of transfers}'' or ``\emph{requirement of Pigou-Dalton}'': transfers from richer households lead to a reduction of inequality. However, several drawbacks are reported in the literature. The most important point is, that the underlying distributional form of the measured inequality is unknown and it is therefore not possible to see if the measure is driven by a few rich or many poor individuals. This can also be problematic for comparison between countries or over time. In extreme cases two totally different distributions share the same Gini-coefficient.\\

@@

\begin{table}[h]
\begin{tabular}{lcccc}
                                   & FTA       & Survey data\\
Statistical unit                   & Tax units (individuals/married couples)  & Households\\                                   
Concepts of economic resources  & Data-driven   & Theory-driven\\
Population coverage & total population & sample \\ 
Top incomes covered                 & Yes   & No\\
Low incomes covered?               & no        & yes\\
Comparison over time possible 	   & Long       & Short \\
Cross country comparison possible? & limited   & possible

\end{tabular}
\end{table}

%RF: damit haben wir WER, WANN, WAS, wollen wir noch ein WOMIT/WAS DARSTELLEN -> methodik -> reldist ? jaa

To define a standard of measuring economic resources and related inequality we discussed different dimensions of demands the data needs to meet. To sum up, ideally we want to measure \emph{disposable income} for \emph{all} swiss \emph{households} yearly for an extended period of time. Tax data has one weakness when it comes to the ideal income measure as we need to work with ``taxable income'' which is designed to serve taxation purposes. However, both concepts -- what needs to be taxed and what can be spent -- are related as both address a distinction between necessary and voluntary expenses. By subtracting taxes from taxable income we are confident to satisfyingly come close the theoretically ideal income measure of disposable income. The second most important drawback with tax data is that it does not adequately address households. There are few situations in which the tax unit equals the household, that is for individuals living alone, with married partner and/or deductible children and nobody else. This ``classic'' household setup however became less common in the last decades so we need to assume statements about household inequality based on taxed data became more and more biased. The swiss law seems to be slightly outdated in this context.
Finally however, another advantage of tax data (over survey data) is indeed the observation period. While the FTA data we use range back to 1945, the earliest period of survey data is 2000 (HBS data). The striking feature of the long FTA time series is its consistency. Both population coverage and income measures are rather consistent for the complete observation period\footnote{Limitations might however exist due to tax evasion and major changes in the tax system, e.g. 1945. Furthermore, tax units with incomes too small to quality for federal taxation are not documented before 1995 and treated as zeros afterwards making it difficult to track income development in the lower percentiles.}. This can not be said for survey data as these are based on samples and therefore require an ideal sampling design or reweighting to be representative and comparable over time. We will see how well sampling and weighting is done in the results section.

@@

\subsection{Outline of applied methods}

@@

In the last section we described the advantages and drawback of taxdata discussing five aspects, which we regard as crucial concering the assessment of inequality. To get a feeling of the importance of these aspects, we exploit the FTA-tax data as far as possible and perform several insightful calculations adressing four of the five mentioned aspects. No further investigation is possible regarding aspect (1) concepts on measuring economic ressource. This is a point which cannot be further addresses with FTA tax data. But for the other four thematic areas, we can provide deeper insight. In general our main strategy is to apply different possible concepts within one conceptional area (income, units, measurement, population coverage) while holding other conceptual differences constant. With this strategy we want to show, the sensitiveness the assessement of inequality is especially if one looks at time trens.  In this section we  describe which methods we use to produce the results in the next section.

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/zero_descriptives}

@@

The number of zeros increased slightly. Different interpretations seems possible. The increased share of zeros could be driven by the FTA's inflation adjustments or an increase in social deductions. Another plausible explanation is the increasing number of unmarried couples filling in two separate tax forms instead of one so we observe two ``small'' incomes instead of one ``large'' income, thus inflating the number of people falling below the taxation threshold.

@@

By adding the zeros back as a separate group we need to make an assumption about the income distribution within that group. We know these cases had incomes between zero and the threshold (usually around 15.000 CHF for the period 1995-2010). In figure WITHWITHOUTZEROSTIMESERIES we show Gini coefficients calculated under the assumption of the zeros being zero. This is problematic for two reasons: first we know these units might earn more (up to the threshold) so Ginis calculated on this assumptions will be too high. Second we assume the zero group to be perfectly homogenous which on the contrary has a lowering effect on the Gini coefficient. Therefore we must at least assume a distribution for the zeros that has some variance and ranges from zero to the threshold. We decided to use the most simple approach fulfilling both criteria which is to interpolate a uniform distribution between zero and each tax periods threshold.

The second approach investigates whether the FTA time series is suitable to analyze trends of inequality even if the inequality measures are biased. We can do this by regressing the inequality measure (here the Gini coefficient) on the share of zeros. Simply spoken we test if cutting a larger (left) tail of the distribution changes the shape of the distribution. We include Gini coefficient for all cantons in our analyses to improve the power of our statistical test\footnote{Furthermore without cantonal data (but only data for aggregate Switzerland) it is impossible to distinguish whether inequality changed due to the number of unobserved people or any other variable that changed over time)}.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/corrected_gini}

@@

\emph{Gini coefficients for taxable income, income before deductions and taxable income after fedaral tax}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/different_ginis}

@@

\emph{Equivalence scale}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
\includegraphics[width=\maxwidth]{figure/equivalencescale}

@@

% Vergleich über die Zeit da, wo zum ersten Mal Gesamtschweizerische Daten vorliegen bis zur neusten Reihe: 2003 vs 2010

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
## Smoothing using 0.4
\end{verbatim}
\end{kframe}
\includegraphics[width=\maxwidth]{figure/reldist20032010}

@@

\begin{kframe}

{\ttfamily\noindent\itshape\color{messagecolor}{\#\# Loading required package: splines}}\end{kframe}% latex.default(format(c(`Median Index` = med, `Lower Index` = lower,      `Upper Index` = upper, `$\\Delta$ Gini` = dgini), digits = ndig),      file = "", title = title) 
%
\begin{table}[!tbp]
\begin{center}
\begin{tabular}{ll}
\hline\hline
\multicolumn{1}{l}{Inequality Indices}&\multicolumn{1}{c}{}\tabularnewline
\hline
Median Index&0.058\tabularnewline
Lower Index&0.072\tabularnewline
Upper Index&0.045\tabularnewline
$\Delta$ Gini&0.025\tabularnewline
\hline
\end{tabular}
\end{center}
\end{table}

@@

% FEHLT: Tabelle mit Median, lower und upper polaritätsindex plus Gini bzw. Differenz der Ginis

@@

\subsubsection{Normal versus special cases}

The FTA stopped to publicly report data for special cases after tax period 1993/94. 
For more recent periods however we can compare already calculated percentiles provided by the FTA\footnote{These calculations were done on commission of the FTA within the SNF project Sinergia Nr. 130648 "The Swiss Confederation: A Natural Laboratory for Research on Fiscal and Political Decentralization" by Raphael Parchet and Stefanie Brilon in coordination with Prof. Dr. Marius Brülhart.} which are based on both special and normal cases to normal cases alone.
We will have a look at both, the 1993/94 period as the last period where both 
numbers were reported as well as 2010 where we compare data from the same source (FTA)
but based on different reporting (aggregated data vs. percentiles).

>>>>

\usepackage[hmarginratio=1:1,top=32mm,columnsep=20pt]{geometry} % Document margins
\usepackage{multicol} % Used for the two-column layout of the document
\usepackage{float} %we need this to use figure environments within multicol. knitr figure chunks need to include fig.pos='H'. 'H' stems from package float and deactivates floating (which multicol does not like)
\usepackage[hang, small,labelfont=bf,up,textfont=it,up]{caption} % Custom captions under/above floats in tables or figures
\usepackage{booktabs} % Horizontal rules in tables
\usepackage{float} % Required for tables and figures in the multi-column environment - they need to be placed in specific locations with the [H] (e.g. \begin{table}[H])
\usepackage{hyperref} % For hyperlinks in the PDF
\usepackage{graphicx}
\usepackage{datetime} %for automatic date in the header

@@

\usepackage{fancyhdr} % Headers and footers
\pagestyle{fancy} % All pages have headers and footers
\fancyhead{} % Blank out the default header
\fancyfoot{} % Blank out the default footer
\newdateformat{mydate}{\monthname[\THEMONTH] \THEYEAR}
\fancyhead[C]{Assessing income inequality with tax data $\bullet$ \mydate\today~$\bullet$ Working paper} % Custom header text
\fancyfoot[RO,LE]{\thepage} % Custom footer text
%\bibliographystyle{plain}
\bibliographystyle{unsrtnat}

@@

% 3.Abschnitt: What do we know about Switzerland
% Offical Data Collections on development of inequality
As we focus our paper on the case of Switzerland, it is important to embed our work in the context of given publications concerning inequality in income. What is known about Switzerland so far? Looking for official data, three main sources has to be mentioned, which can be considered as de facto official data sources: EU-SILC, HBS and LIS-data. Figure \ref{fig:SwissinequalityPlot} shows the results stemming from this three sources while looking at Gini of equivalised disposable income. Up to the day, EU-SILC or Statistics on Income and Living Conditions is the main source used for policy monitoring at EU-level. The main focus of EU-SILC is to collect data on a common ``framework'' to ensure comparability among EU-countries and countries living around or within the EU. As a Non-EU member Switzerland implemented the instrument not from the beginning (2004) but as from 2007. Therefore this times-series doesn't cover time points before 2007. As graph XY shows, following the results from SILC income inequality decreased from 2007 to 2013.\footnote{Data shown in the graph was downloaded from the Eurostat Metadata-portal \url{http://epp.eurostat.ec.europa.eu/cache/ITY_SDDS/EN/ilc_esms.htm} last accessed 21.Mai 2014.} The second important source concerning the distribution of income is the Household Budget Survey (HBS). The main focus of this survey lays in providing detailed data on household budgets. This allows researcher to look at different income concepts like income before and after public transfers. Since 2000 the survey has been conducted on a continuous basis, which allows to look at a consistent time series from 2000 to 2011. As it can be seen from figure \ref{fig:SwissinequalityPlot} the trend is rather stable.\footnote{figures shown in the graph were calculated out of the original datasets, which were kindly provided by Swiss Federal Statistical Office.} Both time-series (SILC and HBS) cover a relatively short time period. A longer period is covered in the LIS-Data-set (1982-2004). Data-provider for the LIS Data is the Swiss Federal Statistical Office too. In contrast to the aforementioned surveys the LIS-data is harmonized out of three surveys: Swiss Income and Wealth Survey (1982), Swiss Poverty Survey (1992) and the Income and Consumption survey (2000, 2002, 2004).  All in all the LIS dataset contains the longest time series on inequality for Switzerland. Analyzing this data \citet{gornick_income_2013} found for Switzerland a quite substantially decreases in income inequality, contradictory to the development in most other western countries. This result is supported by \citet{grabka_evolution_2012} analyzing the Swiss Household Panel (2000-2009).\footnote{A further official database for income distribution is the OECD-Database. It includes measures from Income and Consumption survey as well. Additional data for 2008 is available from EU-Survey of Income and Living Conditions (EU-SILC). But this change in survey is considered as a strict break. Comparison before and since 2008 is not recommended (OECD 2012:315). For the sake of completeness the database construtcted by \citet{deininger_new_1996} and the World Income Inequality Database (WIID) have to be mentioned. Both datasets do not contain figures  for recent years for Switzerland. A further important database on inequality is the GINI database which has been derived from the GINI Country Reports \citep{nolan_changing_2014}. But this dataset doesn't cover Switzerland.}  \\

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}[H]

\includegraphics[width=\maxwidth]{figure/SwissinequalityPlot} \caption[Trends in income inequality]{Trends in income inequality\label{fig:SwissinequalityPlot}}
\end{figure}

@@

\emph{Measuring inequality or concentration} \\
To be able to make qualifying statements about a distribution or to compare different distributions, the concept of inequality turned out to be the most appropriate and thus the most commonly used dimension. The Gini coefficient is the most known measure and mainly used for international comparison. As it is derived from the Lorenz-curve, the quantified amount of inequality can unpretentiously be described in a formal and visual way. Therefore the Gini coefficient is easily interpretable. Furthermore it has several desired statistical properties \citet{engelhardt_modelle_2000}. (1) ``\textit{principle of population}'': the assessment of inequality is independent of the population size (2) ``\textit{Requirement of Bresciani-Turroni}'': the measure is sensitive for changes of income shares, but not for absolute changes (e.g. doubling of all income) (3) ``\textit{weak principle of transfers}'' or ``\textit{requirement of Pigou-Dalton}'': transfers from richer households lead to a reduction of inequality. However, several drawbacks are reported in the literature. The most important point is, that the underlying distributional form of the measured inequality is unknown and it is therefore not possible to see if the measure is driven by a few rich or many poor individuals. This can also be problematic for comparison between countries or over time. In extreme cases two totally different distributions share the same Gini-coefficient.\\

@@

\multicolinterrupt{
\begin{table}[H]
\caption{Comparison of tax and survey data\label{datacomp}} 
\begin{center}
\begin{tabular}{lcc}
\hline\hline
                                   & Tax data       & Survey data\\
\hline
Statistical unit                   & tax units (individuals/married couples)  & households\\                                   
Concepts of economic resources  & data-driven   & theory-driven\\
Population coverage & total population & sample \\ 
Top incomes covered                 & yes   & no\\
Low incomes covered?               & no        & yes\\
Comparison over time possible 	   & long       & short \\
Cross country comparison possible? & limited   & possible\\
\hline
\end{tabular}
\end{center}
\end{table}
}
%RF: damit haben wir WER, WANN, WAS, wollen wir noch ein WOMIT/WAS DARSTELLEN -> methodik -> reldist ? jaa

To define a standard of measuring economic resources and related inequality we discussed different dimensions of demands the data needs to meet. To sum up, ideally we want to measure \textit{disposable income} for \textit{all} swiss \textit{households} yearly for an extended period of time. Tax data has one weakness when it comes to the ideal income measure as we need to work with ``taxable income'' which is designed to serve taxation purposes. However, both concepts -- what needs to be taxed and what can be spent -- are related as both address a distinction between necessary and voluntary expenses. By subtracting federal taxes from taxable income we are getting closer to the theoretically ideal income measure of disposable income. However we can not subtract the more important cantonal and communal taxes. 
The second most important drawback with tax data is that it does not adequately address households. There are few situations in which the tax unit equals the household, that is for individuals living alone, with married partner and/or deductible children and nobody else. This ``classic'' household setup however became less common in the last decades so we need to assume statements about household inequality based on taxed data became more and more biased. 
%The swiss law seems to be slightly outdated in this context.
Finally however, another advantage of tax data (over survey data) is indeed the observation period. While the FTA data we use range back to 1945, the earliest period of survey data is 2000 (HBS data). The striking feature of the long FTA time series is its consistency. Both population coverage and income measures are rather consistent for the complete observation period\footnote{Limitations might however exist due to tax evasion and major changes in the tax system, e.g. 1945. Furthermore, tax units with incomes too small to quality for federal taxation are not documented before 1995 and treated as zeros afterwards making it difficult to track income development in the lower percentiles.}. This can not be said for survey data as these are based on samples and therefore require an ideal sampling design or reweighting to be representative and comparable over time. We will see how well sampling and weighting is done in the results section.

@@

\section{Methods}

@@

In the last section we described the advantages and drawback of tax-data discussing five aspects, which we regard as crucial concerning the assessment of inequality. To get a feeling of the importance of these aspects, we exploit the FTA-tax data as far as possible and perform several insightful calculations addressing four of the five mentioned aspects. No further investigation is possible regarding aspect (1) concepts on measuring economic resource. But for the other four thematic areas, we can provide deeper insight. In general our main strategy is to apply different possible concepts within one conceptional area (income, units, measurement, population coverage) while holding other conceptual differences constant. With this strategy we want to show, the sensitiveness the assessment of inequality is especially if one looks at time trends.\\

To fulfill the above described task, we use two techniques. To assess the development over time, we calculate in general Gini-coefficients for all possible time points, allowing us the make time trends visible and to compare the result out of the FTA-tax data to other publications. For selected periods we expand the analysis with relative distribution methods, which allows an in-depth analysis of distributional differences and therefore catches the shortcomings of Gini-coefficients. \\

To calculate Gini-coefficients we follow common definitions, like it is described for example in \citep{jann_einfuhrung_2005}. Because the relative distribution methods are not that widely known, we provide a short introduction of the important concepts in this section. We additional expand the method section with an extended analysis about the importance of non-taxed, because the issue of non-taxed is a especially thorny one.

\subsection{Relative distribution (RD) framework}

The RD-framework is based on the concept of a ``relative distribution'', a transformation of the data from two distributions into a single distribution that contains all of the information necessary for scale-invariant comparison. This allows to make distribution differences ``visual'' in an elegant way and it is also a base for summary statistics, which are more sensitive to detailed theoretical hypotheses in contrast to other measures like the Gini-coefficient or top income-shares, which inform either about the whole population or only about the top part of the distribution.\\

% Probability Density function as a base for the RD
The goal of RD is to study the differences between two distributions. A common example could be the income distribution for man and women. Subject of the comparison can also be two distributions describing the same population, but stemming from two different data sources like survey or tax data or even comparison of the same source/population but for different time points, like the income distribution out of tax data for Switzerland today compared to an earlier time point. To describe how the two distributions are gone be transformed into a relative distribution, we start with defining the two distributions. One represents the reference population $Y_{0}$ and the other the comparison population $Y$. $x$ represents our measure of interest (income). A first visual approach is to compare the two probability density functions (PDF). The PDF is a function $f(x)$ which describes the distribution of probability over the outcome set and is defined for all possible values of $x$. This function integrates to 1, which means that the sum of all probabilities over all possible values is 100\%. Out of the comparison of the PDF, it is possible to see, which values of $x$ are more and less probable. This already allows to spot distributional differences over the whole scale of $x$ visually. \\

The PDF can be characterized by its cumulative distribution function (CDF). The CDF can be formulated as $F(x)$, which represents the probability that a randomly chosen value is less than or equal to $x$. The relative distribution of $Y$ to $Y_{0}$ is then defined as

\begin{equation}
R=F_{0}(Y)
\end{equation}

$R$ is obtained from $Y$ by transforming it by the CDF for $Y_{0}$, $F_{0}$. $R$ therefore measures the relative rank of $Y$ compared to $Y_{0}$.

% Relative Distribution

\begin{equation}
g(r)=\frac{f(F_{0}^{-1}(r))}{^{f_{0}(F_{0}^{-1}(r))}}
\end{equation}

We can calculate the Probability Density Function $g(r)$ of R, where $r$ represents the proportion of values and $F_{0}^{-1}(r)$ is the inverse cumulative distribution function, also called the quantile function. $g(r)$ can be interpreted as a density ratio, which is defined as the ratio of these two quantities evaluated at every percentile of the reference distribution [0,1]. With a complete overlap of both distributions the probability density function of the $R$ is 1 at every point of the PDF. On the other hand, values higher than 1 represent higher probabilities in the comparison distribution than in the references distribution at this specific point and values lower than 1 respectively represent lower probabilities. It is a proper PDF in the sense that it integrates to 1 over the unit interval.\\

% Median and shape differences
What we got through the above transformation of two distributions is the overall relative probability density. But differences between distributions can be divided into two basic components: changes in location and changes in shape. If the comparative distribution is a simple location-shifted version of the reference distribution, then the difference between the two distributions can be parsimoniously summarized by this shift. Differences that remain after a location adjustment are differences in ``shape'' (scale, skewness and other distributional characteristics). When both types of shifts are operating, or when factors other than scale are changing in the shape component, we need a way to separate out the various effects. If we want to identify the effect of a location shift and separate it from other changes in the distribution, it is necessary to specify what scale this shift operates on. It is possible to adjust distributions by any measure of central tendency. Here we choose a median location adjustment because the median is a natural, robust and scale invariant unit of measurement. Because our interest lies in analyzing distributional differences concerning the degree of inequality, we will focus in the results section on shape differences and look therefore at the relative distribution after the distributions are adjusted for location differences.

% R based Summary Meausres
Distributional polarization is of particular interest in the study of inequality. However, common inequality indicators (for example Gini or Theils index) are not designed to distinguish between growth in the upper and lower tails. Even if the measures register increasing inequality over time, one cannot distinguish a polarization of the distribution (increases in both tails) from upgrading (increases in the upper tail) or downgrading (increases in lower tail). The polarization index developed by \citep{Handock and Morris 1999} addresses this issue, because it is decomposable to distinguish differences in the upper and lower tails.  Because it is based on the relative distribution it provides a simple link between what is observed in the graphical display and what is measured by the numerical summary. \\

The median relative polarization index (MRP) is defined as the mean absolute deviation around the median of the location-adjusted relative distribution, scaled to produce an index that varies between -1 and 1. Given the scaling, a value of zero represents no differences in distributional shape; positive values represent more polarization (increases in the tails of the distribution); and negative values represent less polarization (convergence towards the center of the distribution). The measure catches only differences in distributional shape (not location). And it has several interesting features. MRP can be interpreted in terms of a proportional shift of mass in the distribution from more central to less central values. A value of 0.1, for example, is equivalent to a 10\% population shift from the center of the distribution to the upper and lower quartiles and the MRP is decomposable along the scale of $y$. This makes it possible to compare the contribution of each section of the distribution to the overall polarization. A natural decomposition is the contributions made by components above (upper polarization index, URP) and below (lower polarization index, LRP) the median (of $g(r)$).

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}[H]

\includegraphics[width=\maxwidth]{figure/zero_descriptives} \caption[Share of tax units below federal tax liability threshold]{Share of tax units below federal tax liability threshold\label{fig:zero_descriptives}}
\end{figure}

@@

The number of zeros increased slightly according to figure \ref{fig:zero_descriptives}. Different interpretations seems possible. The increased share of zeros could be driven by the FTA's inflation adjustments or an increase in social deductions. Another plausible explanation is the increasing number of unmarried couples filling in two separate tax forms instead of one so we observe two ``small'' incomes instead of one ``large'' income, thus inflating the number of people falling below the taxation threshold.

@@

By adding the zeros back as a separate group we need to make an assumption about the income distribution within that group. We know these cases had incomes between zero and the threshold (usually around 15.000 CHF for the period 1995-2010). In figure \ref{fig:with_without_zeros_Gini} we show Gini-coefficients calculated under the assumption of the zeros being zero. This is problematic for two reasons: first we know these units might earn more (up to the threshold) so Ginis calculated on this assumptions will be too high. Second we assume the zero group to be perfectly homogeneous which on the contrary has a lowering effect on the Gini coefficient. Therefore we must at least assume a distribution for the zeros that has some variance and ranges from zero to the threshold. We decided to use the most simple approach fulfilling both criteria which is to interpolate a uniform distribution between zero and each tax periods threshold.

The second approach investigates whether the FTA time series is suitable to analyze trends of inequality even if the inequality measures are biased. We can do this by regressing the inequality measure (here the Gini-coefficient) on the share of zeros. Simply spoken we test if cutting a larger (left) tail of the distribution changes the shape of the distribution. We include Gini-coefficient for all cantons in our analyses to improve the power of our statistical test\footnote{Furthermore without cantonal data (but only data for aggregate Switzerland) it is impossible to distinguish whether inequality changed due to the number of unobserved people or any other variable that changed over time)}.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}[H]

\includegraphics[width=\maxwidth]{figure/corrected_gini} \caption[Bias variation by time and cantons]{Bias variation by time and cantons\label{fig:corrected_gini}}
\end{figure}

@@

\emph{Gini coefficients for taxable income, income before deductions and taxable income after federal tax}

We start with the effect of income definition on the assessment of income inequality. As described in the previous section income definitions are either closer to the market outcome or the the income, which shapes the possibility to consume (disposable income). With FTA-Tax data in-between income measures can be defined. This includes

\begin{itemize}
  \item \textit{taxable income}: all incomes declared for taxes (earnings, income from property and current transfers received) minus several deductions
  \item \textit{net income}: Is a sort of total income, which excludes some (but not all) deductions
  \item \textit{taxable income after federal tax}: By subtracting federal taxes from the taxable income, we get closer to the disposable income
\end{itemize}

Following this three definitions we calculate Gini-coefficients out of the FTA-tax statistics. Because information about net income per income bracket is only reported from 1983/1984 until 2010, we can show differences over this time period.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}[H]

\includegraphics[width=\maxwidth]{figure/different_ginis} \caption[Gini coefficients for different income measures]{Gini coefficients for different income measures\label{fig:different_ginis}}
\end{figure}

@@

As \ref{fig:different_ginis} shows, the development for the three defined measures of income is quit parallel except for the 1980s. In this time periode the Gini-coefficent for net income veers. This has probably to do with a change in regulations of deductions and shows that interpretation over time has to be very carefull, becauses changes in taxation or regulation systems can affect the outcome. In generall inequality assesst with taxable income is higher than inequality assesst with net income or taxable income after federal taxes. While it is not surprising that federal taxes reduces inequality slightly because of the progressivity of the taxes and it is somehow astonishing that adding deductions (from net income to taxable income) increases inequality.

\subsection{Statistical units -- Equivalence scale}

The main issue concerning the unit of analysis is not easy solved with tax-data, because the concepts of tax-units and households are not perfectly congruent (we will take up this issue, when comparing survey data and the FTA tax data later on.) But we can examine, how the assessment of inequality is affected by the implementation of equivalence scale. We do this by looking and Gini-time series for net income with and without implementation of a equivalence scale. The scale is constructed by using information out of tax data. The income of single household/tax unit are divided by 1 (no change), for married the equivalence-factor is 1.5. For every child or other by the tax-unit supported person a value of 0.3 is added. This measures are provided directly by the FTA and are not calculated by us. Because excluding the group of not-taxed leads to a longer time-series we provide four time-series in total (two possibilities to compare the effect of a equivalence scale)

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}[H]

\includegraphics[width=\maxwidth]{figure/equivalencescale} \caption[Gini coefficients with and without equivalization]{Gini coefficients with and without equivalization\label{fig:equivalencescale}}
\end{figure}

@@

Here we examine how interpretation can change, when we expand the analysis by using relative distribution methods and not only look at Gini-Coefficients. We therefore use the published percentiles about the distribution of taxable income  on the FTA webpage . We prefer these measures over the calculates measures out of the published income brackets statistics, because they represent the distribution at both tails more accurate since they are based directly on the information about every single tax units\footnote{When calculating percentiles out of the income bracket statistic we loose relevant information at the edges. First, we don't have information about taxable income of tax-units falling below the income threshold for federal taxation. We only know how many persons fall in this category. However, the percentiles reported on the FTA webpage are based on the true taxable income (also for units below the threshold), which allows a more precise estimation of the lower percentiles. Secondly, it is especially hard to estimate the highest top income percentiles out of the aggregated tax statistics, leaving us with information only until the 95\%-percentiles, while the reported percentiles reach the 99.99\%-percentiles}. We use the reported Measures at the cost of time. The longest time-period we can compare out of these data reach from 2003 to 2010. Gini changed during this time from 0.47 to 0.50, which equals a moderate increase of inequality. The in-depth distributional analysis allows us to see, how this change translated into different shapes of the distributions. \\

As described in the method section, the analysis includes a comparison of the density distribution of the log taxable income. We define the 2003 distribution as the reference group and compare it to the distribution of 2010. We then compare the relative density of the median adjusted distributions, to see where the two distributions differ. Finally, we calculate the median relative polarization index (MRP), to quantify the degree of polarization. We also show the decomposition into the upper and lower polarization index (URP and LRP).

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}[H]

\includegraphics[width=\maxwidth]{figure/reldist20032010} \caption[Distribution change over time]{Distribution change over time\label{fig:reldist20032010}}
\end{figure}

@@

%latex.default(format(c(`Median Index` = med, `Lower Index` = lower,     `Upper Index` = upper, `$\\Delta$ Gini` = dgini), digits = ndig),     file = "", title = title, where = "H", label = tablabel,     caption = caption)%
\begin{table}[H]
\caption{\label{pol20032010}} 
\begin{center}
\begin{tabular}{ll}
\hline\hline
\multicolumn{1}{l}{Inequality Indices}&\multicolumn{1}{c}{}\tabularnewline
\hline
Median Index&0.058\tabularnewline
Lower Index&0.072\tabularnewline
Upper Index&0.045\tabularnewline
$\Delta$ Gini&0.025\tabularnewline
\hline
\end{tabular}\end{center}

\end{table}

@@

While looking relative density of shape differences (see right part of \ref{fig:reldist20032010}) it gets clearly visible, how the two compared distributions differ. The pattern suggest that from 2003 to 2010 a moderate polarization occurred, which is represented in a lower relative density in the middle seven deciles (d.30 to d.90), while the density ratio is higher in the top decile and the two lowest deciles. Comparing 2003 to 2010 tax unites moved to the tails. This pattern is quantified with the inequality indices reported in table 2. Comparing the lower and the upper index shows, that the polarization is slightly stronger driven by the downgrading process.

@@

\emph{Normal versus special cases}

The FTA distinguishes normal from special cases as described in the data section. To test whether it matters which cases the researcher looks at we want to compare the distributions of normal and special cases. Unfortunately, the FTA stopped to publicly report data for special cases after tax period 1993/94. Therefore we will compare the two distributions for a rather old dataset. However the FTA does report aggregate statistics (e.g. percentiles) based on a pool of all cases (normal and special) for more recent periods\footnote{These calculations were done on commission of the FTA within the SNF project Sinergia Nr. 130648 "The Swiss Confederation: A Natural Laboratory for Research on Fiscal and Political Decentralization" by Raphael Parchet and Stefanie Brilon in coordination with Prof. Dr. Marius Brülhart.} which allows us to do a corresponding analysis for 2010 as well.
